---
title: "Analysis of Raw GT3X files to Summary Measures in Chadwell et. al Data"
author: "John Muschelli"
date: '`r Sys.Date()`'
output: 
  html_document:
    keep_md: true
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true
bibliography: refs.bib  
editor_options: 
  chunk_output_type: console    
---

All code for this document is located at [here](https://raw.githubusercontent.com/muschellij2/osler/master/gt3x_limb_data/index.R).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, comment = "")
```

```{r packages}
# remotes::install_github("muschellij2/SummarizedActigraphy")
library(SummarizedActigraphy)
# remotes::install_github("THLfi/read.gt3x")
library(read.gt3x)
library(dplyr)
library(readxl)
library(tidyr)
library(readr)
library(lubridate)
library(kableExtra)
library(corrr)
```

# Data

The data is from @chadwell_kenney_granat_thies_galpin_head_2019.  The figshare project for this data is located at https://springernature.figshare.com/collections/Upper_limb_activity_of_twenty_myoelectric_prosthesis_users_and_twenty_healthy_anatomically_intact_adults_/4457855. The gt3x data is located at https://springernature.figshare.com/articles/Unprocessed_raw_30Hz_acceleration_data_stored_as_gt3x/7946189.  

The whole data can be downloaded from [Figshare](https://springernature.figshare.com/articles/Unprocessed_raw_30Hz_acceleration_data_stored_as_gt3x/7946189) directly or can be downloaded for each gt3x file from the [Duplicated Data Figshare]().

## Data Description

The data consists of 40 subjects, 20 with prostheses, 20 without.  Each wore tri-axial Actigraph watches for 7 days, one on each hand.  The metadata/demographics is located at [Demog Figshare]()

```{r auth, include=FALSE}
token_file = here::here("fs_token.rds")
if (file.exists(token_file)) {
  token = readr::read_rds(token_file)
  assign("oauth", token, envir = rfigshare:::FigshareAuthCache)
}
```

```{r, echo = TRUE, eval = FALSE}
data_dir = tempdir()
```

```{r, eval = TRUE, echo = FALSE}
data_dir = here::here("data")
```

```{r get_fs_data}
outfile = here::here("data", "file_info.rds")
if (file.exists(token_file) && !file.exists(outfile)) {
  x = rfigshare::fs_details("11916087")
  
  files = x$files
  files = lapply(files, function(x) {
    as.data.frame(x[c("download_url", "name", "id", "size")],
                  stringsAsFactors = FALSE)
  })
  all_files = dplyr::bind_rows(files)
  readr::write_rds(all_files, outfile)
} else {
  all_files = readr::read_rds(outfile)
}
meta = all_files %>% 
  filter(grepl("Meta", name))
df = all_files %>% 
  filter(grepl("gt3x", name))
df %>% 
  head %>% 
  knitr::kable() %>% 
  kableExtra::kable_styling()
```

We need to add the data path so that it's a full file name:


```{r filedf}
df = df %>% 
  rename(file = name) %>% 
  tidyr::separate(file, into = c("id", "serial", "date"), sep = "_",
                  remove = FALSE) %>% 
  mutate(date = sub(".gt3x.*", "", date)) %>% 
  mutate(date = lubridate::ymd(date)) %>% 
  mutate(group = ifelse(grepl("^PU", basename(file)), 
                        "group_with_prosthesis",
                        "group_without_prosthesis")) %>% 
  mutate(article_id = basename(download_url)) %>% 
  mutate(outfile = file.path(data_dir, group, basename(file)))
df %>% 
  head() %>% 
  knitr::kable() %>% 
  kableExtra::kable_styling()
```


## Demographics data

```{r meta}
metadata = file.path(data_dir, "Metadata.xlsx")
if (!file.exists(metadata)) {
  out = download.file(meta$download_url, destfile = metadata)
}
meta = readxl::read_excel(metadata)
bad_col = grepl("^\\.\\.", colnames(meta))
colnames(meta)[bad_col] = NA
potential_headers = rbind(colnames(meta), meta[1:2, ])
potential_headers = apply(potential_headers, 2, function(x) {
  x = paste(na.omit(x), collapse = "")
  x = sub(" .csv", ".csv", x)
  x = sub(" .wav", ".wav", x)
  x = gsub(" ", "_", x)
  x
})
colnames(meta) = potential_headers
meta = meta[-c(1:2),]
meta = meta %>% 
  rename(id = Participant_Identifier)
meta = meta %>% 
  filter(!is.na(id),
         id != "Participant Identifier") %>% 
  mutate_all(.funs = function(x) gsub("ü", "yes", x))
meta = meta %>% 
  mutate_at(
    .vars = vars(
      Age,
      `Time_since_prescription_of_a_myoelectric_prosthesis_(years)`
    ),
    readr::parse_number
  ) 
meta = meta %>% 
  mutate( `Time_since_limb_loss_(years)` = ifelse(
    `Time_since_limb_loss_(years)` == "Congenital", 0,
    `Time_since_limb_loss_(years)`)
  )
meta %>% 
  head %>% 
  knitr::kable() %>% 
  kableExtra::kable_styling()
```

# Packages

THe `read.gt3x` and `AGread` packages can read gt3x files, but only the `read.gt3x` package can read in the old GT3X format from NHANES 2003-2006.  Thus, we will use that package to read the `gt3x` format.  If you need additional information, such as temperature, lux, etc, you may want to try `AGread::read_gt3x`.  Additionally, the `read.gt3x` package can read in `gt3x` files that have been zipped, including gzipped (extension `gz`), bzipped (`bz2`), or xzipped (`xz`).

THe `SummarizedActigraphy::read_actigraphy` wraps the `read.gt3x` functionality, and puts the output format to the `AccData` format, which is common in the `GGIR` package.  The `read_actigraphy` also tries to read other formats, by using `GGIR::g.readaccfile` and other `GGIR` functionality.  


```{r merge}
data = full_join(meta, df)
```

## Read in one file


Here we will read in one file:

```{r readin}
iid = sample(nrow(df), 1)
idf = df[iid, ]
if (!file.exists(idf$outfile)) {
  out = curl::curl_download(idf$download_url, destfile = idf$outfile)
}
# acc = read.gt3x(idf$outfile, verbose = FALSE, 
#                 asDataFrame = TRUE, imputeZeroes = TRUE)
acc = read_actigraphy(idf$outfile, verbose = FALSE)
head(acc$data.out)
options(digits.secs = 2)
head(acc$data.out)
acc$freq
```

Let's look at the number of measurements per second to see the true sampling rate:

```{r sample_size, dependson="readin"}
res = acc$data.out %>% 
  mutate(dt = floor_date(time, "seconds")) %>% 
  group_by(dt) %>% 
  count()
table(res$n)
```

```{r p_diff, dependson="readin"}
library(ggplot2)
res = acc$data.out %>% 
  mutate(day = floor_date(time, "day"),
         time = hms::as_hms(time)) %>% 
  mutate(day = difftime(day, day[1], units = "days")) 
res = res %>%
  filter(day == 1)
res = res %>% 
  filter(between(time, 
                 hms::as_hms("10:00:00"),
                 hms::as_hms("10:30:00"))
         ) 
res = res %>% 
  tidyr::gather(key = direction, value = accel, -time, -day)

res %>%
  ggplot(aes(x = time, y = accel, colour = direction)) +
  geom_line() +
  theme(
  legend.background = element_rect(
    fill = "transparent"),
  legend.position = c(0.5, 0.9),
  legend.direction = "horizontal",
  legend.key = element_rect(fill = "transparent", 
                            color = "transparent") ) 
  
```

```{r}
check_zeros = function(df) {
  any(rowSums(df[, c("X", "Y", "Z")] == 0) == 3)
}
fix_zeros = function(df, fill_in = TRUE) {
  zero = rowSums(df[, c("X", "Y", "Z")] == 0) == 3
  names(zero) = NULL
  df$X[zero] = NA
  df$Y[zero] = NA
  df$Z[zero] = NA
  if (fill_in) {
    df$X = zoo::na.locf(df$X, na.rm = FALSE)
    df$Y = zoo::na.locf(df$Y, na.rm = FALSE)
    df$Z = zoo::na.locf(df$Z, na.rm = FALSE)
    
    df$X[ is.na(df$X)] = 0
    df$Y[ is.na(df$Y)] = 0
    df$Z[ is.na(df$Z)] = 0
  }
  df
}
```


Here we can do a simple data check:
```{r header}
acc$header
acc$header %>% 
  filter(Field %in% c("Sex", "Age", "Side"))
data[iid, c("Gender", "Age")]
```

```{r create_functions}
calculate_ai = function(df, epoch = "1 min") {
  sec_df = df %>% 
    mutate(
      HEADER_TIME_STAMP = lubridate::floor_date(HEADER_TIME_STAMP, "1 sec")) %>% 
    group_by(HEADER_TIME_STAMP) %>% 
    summarise(
      AI = sqrt((var(X) + var(Y) + var(Z))/3),
    )
  sec_df %>% mutate(
    HEADER_TIME_STAMP = lubridate::floor_date(HEADER_TIME_STAMP, epoch)) %>% 
    group_by(HEADER_TIME_STAMP) %>% 
    summarise(
      AI = sum(AI)
    )
}

calculate_mad = function(df, epoch = "1 min") {
  df %>% 
    mutate(         
      r = sqrt(X^2+Y^2+Z^2),
      HEADER_TIME_STAMP = lubridate::floor_date(HEADER_TIME_STAMP, epoch)) %>% 
    group_by(HEADER_TIME_STAMP) %>% 
    summarise(
      SD = sd(r),
      MAD = mean(abs(r - mean(r))),
      MEDAD = median(abs(r - mean(r)))
    )
}

calculate_measures = function(df, epoch = "1 min") {
  ai0 = calculate_ai(df, epoch = epoch)
  mad = calculate_mad(df, epoch = epoch)
  res = full_join(ai0, mad)
  res
}

```

As the `imputeZeros` function in `read.gt3x` puts zeros for the idle sleep mode in ActiGraph, we need to repeat the measure to mimic the ActiGraph software:

```{r zeroes}
df = acc$data.out
df = df %>% 
  rename(HEADER_TIME_STAMP = time) %>% 
  select(HEADER_TIME_STAMP, X, Y, Z)
zero_rows = rowSums(df[, c("X", "Y", "Z")] == 0) == 3
any(zero_rows)
ind = unname(which(zero_rows))
ind = head(ind)
ind = c(min(ind) - (1:5), ind)
as.data.frame(df[ind,])
df = fix_zeros(df)
as.data.frame(df[ind,])
```
We may want to remove these rows, but we're trying to mimic ActiGraph output.  Now we can calculate a set of "measures" that are estimates of activity.  Those include AI, MAD, and SD.  I have re-implemented the method from the `ActivityIndex` package, with the restriction that no estimate of $\sigma_0$ is given, which is the estimate of the standard deviation when the device is at rest. We see, even for a relatively large number of values, it is pretty quick to compute:

```{r make_measures, dependson="zeroes"}
system.time({measures = calculate_measures(df)})
```

We will calculate MIMS units with the `MIMSunit` package:

```{r MIMS, dependson="zeroes"}
library(MIMSunit)
hdr = acc$header %>% 
  filter(Field %in% c("Acceleration Min", "Acceleration Max")) %>% 
  mutate(Value = as.numeric(Value))
dynamic_range = range(hdr$Value)
system.time({
  mims = df %>% 
    mims_unit(epoch = "1 min", 
              dynamic_range = dynamic_range)
})
measures = full_join(measures, mims)
```


We can show the correlation of the measures with the others, noting that some of these have very high correlation.

```{r corr, dependson="zeroes"}
library(corrr)
measures %>% 
  select(-HEADER_TIME_STAMP) %>% 
  correlate() %>% 
  stretch(remove.dups = TRUE, na.rm = TRUE) %>% 
  arrange(desc(r))
```

Now we can create an average day profile.  We will calculate the mean value of these functions for each minute separately:
```{r avg}
to_minute = function(x) {
  x = format(x, "%H:%M:%S")
  x = hms::as_hms(x)
  x
}
average_day = measures %>% 
  mutate(HEADER_TIME_STAMP = to_minute(HEADER_TIME_STAMP)) %>% 
  group_by(HEADER_TIME_STAMP) %>% 
  summarise_at(vars(AI, SD, MAD, MEDAD), mean, na.rm = TRUE)
average_day %>%
  ggplot(aes(x = HEADER_TIME_STAMP, y = AI)) +
  geom_line()

average_day %>%
  ggplot(aes(x = HEADER_TIME_STAMP, y = MAD)) +
  geom_line()
```

We can also make the data 1440 format:

```{r make1440}
measures1440 = measures %>% 
  select(HEADER_TIME_STAMP, AI) %>% 
  mutate(
    date = lubridate::as_date(HEADER_TIME_STAMP),
    HEADER_TIME_STAMP = to_minute(HEADER_TIME_STAMP)) %>% 
  mutate(HEADER_TIME_STAMP = sprintf("MIN_%04.0f", as.numeric(HEADER_TIME_STAMP)/60)) %>% 
  spread(HEADER_TIME_STAMP, value = AI)
head(measures1440)
```





# References
