---
title: "Estimating Thresolds Compared to Vector Magnitude Counts"
author: "John Muschelli"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

All data/code is located at https://github.com/muschellij2/upper_limb_gt3x_prosthesis.

# Setup
Load some packages
```{r load_packages, cache = FALSE}
library(dplyr)
library(readr)
library(here)
library(mgcv)
library(ggplot2)
```

# Reading in Data

Here we load the full data, of all participants for all days, at a minute level:
```{r}
full_file = here::here("data", "all_minute_data.rds")

data = read_rds(full_file) %>% 
  rename(vectormagnitude = `Vector Magnitude`)

data = data %>% 
  mutate(MIMS_UNIT = ifelse(MIMS_UNIT < 0, 0, MIMS_UNIT))
head(data)
```
# Exploration

## MIMS units
```{r mims_plot}
data %>% 
  ggplot(aes(x = vectormagnitude, y = MIMS_UNIT)) + 
  geom_hex()

data %>% 
  ggplot(aes(x = vectormagnitude, y = MIMS_UNIT)) + 
  geom_point(alpha = 0.1)

data %>% 
  filter(MIMS_UNIT > 0) %>% 
  ggplot(aes(x = vectormagnitude, y = MIMS_UNIT)) + 
  geom_point(alpha = 0.1)
```

## AI
```{r ai_plot}
data %>% 
  ggplot(aes(x = vectormagnitude, y = AI)) + 
  geom_hex()
data %>% 
  ggplot(aes(x = vectormagnitude, y = AI)) + 
  geom_point(alpha = 0.1) + geom_smooth(se = FALSE)

data %>% 
  filter(AI > 0) %>% 
  ggplot(aes(x = vectormagnitude, y = MIMS_UNIT)) + 
  geom_point(alpha = 0.1)
```

# Estimation

## MIMS Unit
Here we take some thresholds for activity established for vector magnitude and estimating those thresholds for MIMS units using a generalized additive model (GAM) from the `mgcv` package:
```{r mims_pred}
pred_df = data.frame(vectormagnitude = c(100, 1853, 1952, 2690))

mod = gam(MIMS_UNIT ~ s(vectormagnitude, bs = "cr"), data = data)
predict(mod, newdata = pred_df)
```

We can do the same thing for where vector magnitude is greater than 0:
```{r mims_pred0}
mod_over0 = gam(MIMS_UNIT ~ s(vectormagnitude, bs = "cr"), 
                data = data %>% 
                  filter(vectormagnitude > 0))
predict(mod_over0, newdata = pred_df)
```

## AI

We will do the same thing using activity index:
```{r ai_pred}
mod = gam(AI ~ s(vectormagnitude, bs = "cr"), data = data)

predict(mod, newdata = pred_df)
```

We can do the same thing for where vector magnitude is greater than 0:
```{r ai_pred0}
mod_over0 = gam(AI ~ s(vectormagnitude, bs = "cr"), 
                data = data %>% 
                  filter(vectormagnitude > 0))
predict(mod_over0, newdata = pred_df)
```

## Using Bivariate Estimates

We can use a very simple, but non-parametric estimate of this relationship by using a bivariate CDF estimator.  We round the data because this will have as many unique values as in the observed bivariate distribution:

```{r bivariate}
data = data %>% 
  mutate(
    MIMS_UNIT = round(MIMS_UNIT, 2),
         vectormagnitude = round(vectormagnitude, 1))
bivar = data %>% 
  select(MIMS_UNIT, vectormagnitude) %>% 
  group_by(MIMS_UNIT, vectormagnitude) %>% 
  summarise(n = n(), .groups ="drop")
head(bivar)
head(
  bivar %>% 
  filter(vectormagnitude >= 1851 & vectormagnitude <= 1853)
)
bivar %>% 
  filter(vectormagnitude >= 1851 & vectormagnitude <= 1853) %>% 
  summarise(sum(n *MIMS_UNIT)/sum(n))
bivar %>% 
  filter(vectormagnitude >= 1851.5 & vectormagnitude <= 1852.5) %>% 
  summarise(sum(n *MIMS_UNIT)/sum(n))
```
